---
title: "Run Population Growth Rate models for UK subset using full variable set"
output: 
  html_notebook: 
    fig_height: 15
    fig_width: 15
---
Run additional set of models for UK only, using both ECAD and UKCP09 datasets, to enable comparison between the two (#species=80).

```{r}
## Housekeeping
library(dplyr); library(broom); library(lme4); library(tidyr); library(lmerTest);
library(scales)
```

# Read in base datasets and merge 
All fairly straight-forward, save for the step to ensure that monthly weather variables in the ECAD dataset are correctly aligned with the observations in the bird count dataset. I want monthly temperature and precipitation variables that cover the period from the first count ($N_t$) through to the second count ($N_{t+1}$). Counts are made in approximately April, and therefore the months of interest run from May in year $t$ through to April in time $t+1$. 

Sections in alignment step:

(1) take just the first 1:4 months (filtering out the rest), and update the indexing to run 13:16

(2) because these are now indexed counting from the year prior, set the year to be the year prior

(3) merge back to the full ECAD dataframe- note there are now duplicate observations for months 1:4 (but duplicates have 13:16 indexing), but these 1:4 months are not used in subsequent analysis

(4) the counts in the bird dataset are indexed as observation$_{year}$ (count_t) and observation$_{year-1}$ (count_pre); the merging of the ECAD df and the count df is based on the match of the year column. In order for this match to correspond with the period from year-1 to year, indexing of the ECAD df needs to be updated by +1. As an example if you had year==2003 in the ECAD df, and you did a straight merge, all the months of climate data would correspond to the period year:(year+1). By adding 1 to the index, year==2003, becomes year==2004, and the matched climate periods now correspond to (year-1):year, as they should. 

..after this the aligned ECAD datasets are converted to wide-format (ECAD_wide)

## format UKCP dataset
```{r formatUKCP}
UKCP <- readRDS("../files/UKCP09_monthlySumm_UK monitoring sites.rds") %>% 
    ungroup %>%
    mutate(year=as.numeric(year))
# use climate data key to generate analysis grid, as before
climKey <- readRDS("../files/climateKey_siteID-climateID.rds") %>%
    mutate(gridID_200k = paste0("lon_", ceiling(lon_ID/8), ":lat_", ceiling(lat_ID/8)))

# alignment step (see notebook text above)
UKCP_aligned <- UKCP %>% 
    ungroup %>%
    filter(month %in% 1:4) %>%
    mutate(month = month + 12, year = year-1) %>%
    bind_rows(UKCP, .) %>%
    ungroup %>%
    mutate(year = year + 1)

## make wide format
# create wide format df
temp_wide <- UKCP_aligned %>% 
    ungroup %>% 
    dplyr::select(-precip_monthlySum) %>%
    spread(key = month, value=temp_monthlyMean, sep = "_")
names(temp_wide) <- gsub("month", "temp", names(temp_wide))

precip_wide <- UKCP_aligned %>% 
    ungroup %>% 
    dplyr::select(-temp_monthlyMean) %>%
    spread(key = month, value=precip_monthlySum, sep = "_")
names(precip_wide) <- gsub("month", "precip", names(precip_wide))

UKCP_wide <- left_join(temp_wide, precip_wide, by=c("siteID", "year"))
# clear away redundant objects to save some memory
rm(precip_wide, temp_wide, UKCP_aligned)
```

```{r formatECAD}
# Read in ECAD dataset & join in climate cell info from climateKey
ECAD <- readRDS("../files/climateDF_monthlySummary.rds")
# Add 200k analysis grid cell info to climate key (note: climate cells
# are at 25km resolution, so this corresponds to 8 of these)
# climKey <- readRDS("../files/climateKey_siteID-climateID.rds") %>%
#     mutate(gridID_200k = paste0("lon_", ceiling(lon_ID/8), ":lat_", ceiling(lat_ID/8)))
ECAD_full <- left_join(climKey, ECAD, by=c("climateID"))

# alignment step (see notebook text above)
ECAD_aligned <- ECAD %>% 
    ungroup %>%
    filter(month %in% 1:4) %>%
    mutate(month = month + 12, year = year-1) %>%
    bind_rows(ECAD, .) %>%
    ungroup %>%
    mutate(year = year + 1)

## make wide format
# create wide format df
temp_wide <- ECAD_aligned %>% 
    ungroup %>% 
    dplyr::select(-precip_monthlySum) %>%
    spread(key = month, value=temp_monthlyMean, sep = "_")
names(temp_wide) <- gsub("month", "temp", names(temp_wide))

precip_wide <- ECAD_aligned %>% 
    ungroup %>% 
    dplyr::select(-temp_monthlyMean) %>%
    spread(key = month, value=precip_monthlySum, sep = "_")
names(precip_wide) <- gsub("month", "precip", names(precip_wide))

ECAD_wide <- left_join(temp_wide, precip_wide, by=c("climateID", "year"))
# clear away redundant objects to save some memory
rm(ECAD_aligned, ECAD_full, ECAD)
```

# Format bird dataset
```{r formatBirdData}
bird_all <- readRDS("../files/birdDF.rds") 
# create full data frame for analysis
bird_gridded <- left_join(bird_all, dplyr::select(climKey, siteID, gridID_200k, lon_ID, lat_ID), by="siteID") %>%
    filter(schemeID == "UKI.0") %>%
    group_by(species, gridID_200k) %>%
    mutate(n_200k = n())

# filter out grid-cells with fewer than 500 dp & merge in climate data from ECAD
bird_ECAD <- filter(bird_gridded, n_200k >= 500) %>%
    left_join(., ECAD_wide, by=c("climateID", "year"))  

bird_UKCP <- filter(bird_gridded, n_200k >= 500) %>%
    left_join(., UKCP_wide, by=c("siteID", "year"))  

# create df of species to run models for & add columns to catch sigmas
modSumm_toRun <- bird_UKCP %>%
    group_by(species) %>%
    summarise(n_grid=length(unique(gridID_200k)), n_obs = n(), n_scheme=length(unique(schemeID))) %>%
    arrange(n_grid, n_obs)
modSumm_toRun$sigma_null <- NA
modSumm_toRun$sigma_full<- NA
```

# Running models
Iterate across species, fitting models to whichever grid-cells have $\geq$ 500 datapoints. 

First, set up model formulas (`f_full` & `f_null`)
```{r setupModels}
# formula for full model 
varList <- c("temp_5", "temp_6" , "temp_7" , "temp_8" , "temp_9" , "temp_10" , 
             "temp_11" , "temp_12" , "temp_13" , "temp_14" , "temp_15" , "temp_16" , 
             "precip_5" , "precip_6" , "precip_7", "precip_8" , "precip_9" , 
             "precip_10" , "precip_11" , "precip_12" , "precip_13" , "precip_14" , 
             "precip_15","precip_16")

form_full <- paste0("rgr ~ ", paste0(varList, ":gridID_200k", collapse=" + "), 
                 " + logCount_pre:gridID_200k + (1|siteID) + gridID_200k")
form_null <- "rgr ~ logCount_pre:gridID_200k + (1|siteID) + gridID_200k"
print(form_full)
print(form_null)
```

Run models, catching the model coefficients, spatial MSE, and residual variation terms
```{r runModels_UKCP}
# catch model coefficients
coefs <- rep(NA, nrow(modSumm_toRun)) %>% as.list()

for(i in 1:nrow(modSumm_toRun)) {
    sp_i <- modSumm_toRun$species[i]
    print(paste0(i, ":", sp_i))
    df_i <- bird_UKCP %>% filter(species == sp_i)
    
    # scale for numerical stability
    scaledVars <- df_i %>% ungroup %>%
        dplyr::select(temp_1:precip_16, logCount_pre) %>%
        scale(., center = T, scale = T)
    df_i <- bind_cols(df_i %>%  dplyr::select(-c(temp_1:precip_16, logCount_pre)), data.frame(scaledVars))
    
   # run models
    if(modSumm_toRun$n_grid[i] > 1) {
        f_full <- lmer(form_full, df_i)
        f_null <- lmer(form_null, df_i)   
    } else{
        f_full <- lmer(gsub(":gridID_200k| \\+ gridID_200k", "", form_full), df_i)
        f_null <- lmer(gsub(":gridID_200k| \\+ gridID_200k", "", form_null), df_i)  
    }
    
    coef_i <- summary(f_full) %>%
        .$coefficients %>%
        as.data.frame %>%
        mutate(term=row.names(.),
               species = sp_i) %>%
        as_data_frame
    
    coefs[[i]] <- coef_i
    modSumm_toRun[i,"sigma_null"] <- sigma(f_null)
    modSumm_toRun[i,"sigma_full"] <- sigma(f_full)
    
    # diagnostic plot
    png(paste0("../figures/diagnostics/UK/", gsub("/", "-", sp_i),"_UKCP.png"), width = 500, height=500)
    par(mfrow=c(2,2), mai=c(.4, .4, .4,.1))
    qqnorm(residuals(f_full, scale=T), main="full"); abline(0,1)
    qqnorm(residuals(f_null, scale=T), main="null"); abline(0,1)
    plot(residuals(f_full, scale=T)~predict(f_full), col=alpha("black", .5)); abline(0, 0)
    plot(residuals(f_null, scale=T)~predict(f_null), col=alpha("black", .5)); abline(0, 0)
    dev.off()
}
```

```{r format&saveOutputs_UKCP}
modSumm_toRun <- modSumm_toRun %>%
    mutate(var_quotient = sigma_full^2/sigma_null^2)

saveRDS(modSumm_toRun, "../files/model outputs/UKModels_UKCP_explanatoryCapacity.rds")

coefs_full <- bind_rows(coefs) %>%
    as_data_frame()  %>%
    mutate(type=ifelse(grepl("precip", term), "precip",
                             ifelse(grepl("temp", term), "temp", NA))) %>%
    rename(estimate = Estimate, SE = `Std. Error`, tVal = `t value`, pVal = `Pr(>|t|)`)
saveRDS(coefs_full, "../files/model outputs/UKModels_UKCP_coefs.rds")
```


Now repeat, but using ECAD 
```{r runModels_ECAD}
# catch model coefficients
coefs <- rep(NA, nrow(modSumm_toRun)) %>% as.list()

for(i in 1:nrow(modSumm_toRun)) {
    sp_i <- modSumm_toRun$species[i]
    print(paste0(i, ":", sp_i))
    df_i <- bird_ECAD %>% filter(species == sp_i)
    
    # scale for numerical stability
    scaledVars <- df_i %>% ungroup %>%
        dplyr::select(temp_1:precip_16, logCount_pre) %>%
        scale(., center = T, scale = T)
    df_i <- bind_cols(df_i %>%  dplyr::select(-c(temp_1:precip_16, logCount_pre)), data.frame(scaledVars))
    
   # run models
    if(modSumm_toRun$n_grid[i] > 1) {
        f_full <- lmer(form_full, df_i)
        f_null <- lmer(form_null, df_i)   
    } else{
        f_full <- lmer(gsub(":gridID_200k| \\+ gridID_200k", "", form_full), df_i)
        f_null <- lmer(gsub(":gridID_200k| \\+ gridID_200k", "", form_null), df_i)  
    }
    
    coef_i <- summary(f_full) %>%
        .$coefficients %>%
        as.data.frame %>%
        mutate(term=row.names(.),
               species = sp_i) %>%
        as_data_frame
    
    coefs[[i]] <- coef_i
    modSumm_toRun[i,"sigma_null"] <- sigma(f_null)
    modSumm_toRun[i,"sigma_full"] <- sigma(f_full)
    
    # diagnostic plot
    png(paste0("../figures/diagnostics/UK/", gsub("/", "-", sp_i),"_UKCP.png"), width = 500, height=500)
    par(mfrow=c(2,2), mai=c(.4, .4, .4,.1))
    qqnorm(residuals(f_full, scale=T), main="full"); abline(0,1)
    qqnorm(residuals(f_null, scale=T), main="null"); abline(0,1)
    plot(residuals(f_full, scale=T)~predict(f_full), col=alpha("black", .5)); abline(0, 0)
    plot(residuals(f_null, scale=T)~predict(f_null), col=alpha("black", .5)); abline(0, 0)
    dev.off()
}
```


```{r format&saveOutputs_ECAD}
modSumm_toRun <- modSumm_toRun %>%
    mutate(var_quotient = sigma_full^2/sigma_null^2)

saveRDS(modSumm_toRun, "../files/model outputs/UKModels_ECAD_explanatoryCapacity.rds")

coefs_full <- bind_rows(coefs) %>%
    as_data_frame()  %>%
    mutate(type=ifelse(grepl("precip", term), "precip",
                             ifelse(grepl("temp", term), "temp", NA))) %>%
    rename(estimate = Estimate, SE = `Std. Error`, tVal = `t value`, pVal = `Pr(>|t|)`)
saveRDS(coefs_full, "../files/model outputs/UKModels_ECAD_coefs.rds")
```